---
title: "2021 Stream selections"
author: "C Sedlacek"
date: "02/10/2021"
Updated: "09/08/2021 by Jay Silvanima"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rmarkdown)
```

## Contact:
### James (Jay) Silvanima
Department of Environmental Protection

Watershed Monitoring and Data Management

2600 Blair Stone Road

Tallahassee, Florida 32399-2400

James.Silvanima@dep.state.fl.us

phone:   850-245-8507


or

### Andy Woeber

GIS Analyst

Environmental Consultant

Florida Department of Environmental Protection (FDEP)

Division of Environmental Assessment and Restoration (DEAR) 

Watershed Monitoring Section (WMS)

2600 Blair Stone Road, Tallahassee, FL 32399 

850-245-8031

Nathan.Woeber@dep.state.fl.us	



# __Description of Sample Design for Panel Design__
Target population:  All small streams designated as waters of the state of Florida separated into the six reporting zones / basins. 

## __Geographic Information System Sample Frame:__ 
The frame was provided by Florida Department of Environmental Protection, named  _‘Cycle15_SmallStreams_coverage.shp’_.  The length of the streams was calculated based on the length delineated by the GIS coverage.

## __Cycle 15 basins:__

```{r}
# File: Stream selection script.R
# Purpose: Select stream sites for Florida streams
#                 throughout the 6 basins for 2021
# Programmer: Chris Sedlacek, Jay Silvanima
# Date: 25 February 2021
# Updated Sept 2021 to add comments and graphic

# This script needs to be run in R version 3.4 or lower as the grts function
#   will crash otherwise.  Can toggle to 3.4 if you have it loaded via RStudio 
#   Tools global options R versions - change.

##Set directory. This is where the outputs will be saved. 
#   Alter to desired location. Use getwd() to determine the directory for
#     your r project.

setwd("C:/R/2021 Stream selections")

# Load the library
library (spsurvey)

# Load extent of small stream resource coverage in .dbf format from ArcGIS 
# shapefile. Note shape file needs to contain the information for the stream reachcodes and
# and geographic strata (6 Zones).

# Read the .dbf file
att<-read.dbf("C:/R/2021 Stream selections/Cycle15_SmallStreams_Coverage")
names(att)<-tolower(names(att)) # This action just makes typing easier

# Check the first 6 and last 6 lines of data in att
head(att)
tail(att)
levels(att$c3_zone)
#levels(att$stratum)
#levels(att$basin_name)
att$basin_name<- att$c3_zone

# Determine the extent of the streams in km for each zone and for the state.

strmlngth<-tapply(att$length_km,list(att$c3_zone), sum)
strmlngth[is.na(strmlngth)] <- 0
round(addmargins(strmlngth),1)

```
### Zone   Stream length in km

Zone 1 = 12,329.2

Zone 2 = 2,201.8

Zone 3 = 4,484.5

Zone 4 = 4,199.1

Zone 5 = 9,82.6

Zone 6  = 182.5 

Sum = 24,379.8
              
  
  
### __Survey Design:__ 
A Generalized Random Tessellation Stratified (GRTS) survey design for a linear stream resource was used.  The GRTS design includes reverse hierarchical ordering of the selected sites.

First crate a column to use as a stratum variable and check the components in it.  Then set up the panel design.

```{r}

att$stratum<-factor(as.character(att$c3_zone))

levels(att$stratum)
levels(att$c3_zone)

# Create the panel design for the six basins created for 2021 streams 
# selection with a 9x oversample
# Stratified           

dsgn_SS <- list('Zone 1'=list(panel=c(Base=15),
                              seltype='Equal',
                              over=135), 
                'Zone 2'=list(panel=c(Base=15),
                              seltype='Equal',
                              over=135),
                'Zone 3'=list(panel=c(Base=15),
                              seltype='Equal',
                              over=135),
                'Zone 4'=list(panel=c(Base=15),
                              seltype='Equal',
                              over=135),
                'Zone 5'=list(panel=c(Base=15),
                              seltype='Equal',
                              over=135),        
                'Zone 6'=list(panel=c(Base=15),
                              seltype='Equal',
                              over=135)
                
)

```

# Create the GRTS survey design.

First generate a random number and then include in set.seed command.  Then set up spsuvey arguments and run grts function to produce the shape file with the selected site locations. Note there are lots of arguments for the grts function.  type.frame = finite, linear, or area.  scr.frame = source type of target population geographic coverage.  in.shape = shapefile name of coverage.  att.frame = name of dataframe to be created from shapefile.  stratum = column in dataframe containing strata for design.  prjfilename = ArcGIS project file name.  out.shape = loation and name of shapefile to be generated which contains the site selections.

```{r}
sample(1000000,1) 
# run once to get random seed and put result into set.seed
# Reason is so that can reproduce exactly same sites if rerun it.
set.seed(356356) # Don't change unless want a different set of sites
dsgntime <- proc.time()  # keep track of how long spsurvey takes
sites <- grts(design=dsgn_SS,
              DesignID="FLSS21001",
              type.frame="linear",
              src.frame="shapefile",
              in.shape="C:/R/2021 Stream selections/Cycle15_SmallStreams_coverage",
              att.frame=att,
              stratum="stratum",
              prjfilename="C:/R/2021 Stream selections/Cycle15_SmallStreams_coverage",
              out.shape="C:/R/2021 Stream selections/ss site plot")
dsgntime <- (proc.time() - dsgntime)/60
dsgntime

# print summary of sites selected
dsgnsum(sites)
# Print the initial six lines of the survey design
sites <- sites@data    # this is attribute part of shapefile created
head(sites)

write.csv(sites, "2021_Stream_Selections.csv")
```

# plot sites
library(rgdal)

sssiteplot <- readOGR("C:/R/2021 Stream selections/ss site plot.shp")

plot(sssiteplot)
```

### __Stratification:__  
Stratification was by zones/basins.

### __Expected sample size:__  
15 sites within each of the state’s zones/basins. 

### __Oversample:__ 
135 sites for each Cycle 15 basin.

### __Site Use:__  
The base design has 10 sites for each basin in the stratum.  Sites are listed in SiteID order and must be used in that order.  All sites that occur prior to the last site used must have been evaluated for use and then either sampled or the reason documented as to why that site was not used.  As an example, if 30 sites are to be sampled in the watershed, then the first 30 sites in SiteID order would be used.

### __Sample Frame Summary__

Design Summary: Number of Sites Classified by panel and stratum

          stratum
panel      Zone 1 Zone 2 Zone 3 Zone 4 Zone 5 Zone 6 Sum
  Base         15     15     15     15     15     15  90
  OverSamp    135    135    135    135    135    135 810
  Sum         150    150    150    150    150    150 900

```{r}
#Design Summary: Number of Sites Classified by panel and stratum

#          stratum
#panel      Zone 1 Zone 2 Zone 3 Zone 4 Zone 5 Zone 6 Sum
#  Base         15     15     15     15     15     15  90
#  OverSamp    135    135    135    135    135    135 810
#  Sum         150    150    150    150    150    150 900


# The printout above was created with the following code. As the versions update and change it is hoped that the code will function appropriately in the future.
# print summary of sites selected
#dsgnsum(sites)
# Print the initial six lines of the survey design
#sites <- sites@data    # this is attribute part of shapefile created
#head(sites)

```


### __Site Selection Summary__
No output from R.

Description of Sample Design Output:
The sites are provided as a shapefile that can be read directly by ArcMap.  The dbf file associated with the shapefile may be read by Excel.

The dbf file has the following variable definitions:
Variable Name	Description
SiteID	Unique site identification (character)
arcid	Internal identification number
xcoord	Albers x-coordinate
ycoord	Albers y-coordinate
mdcaty	Multi-density categories used for unequal probability selection
weight	Weight (in meters), inverse of inclusion probability, to be used in statistical analyses
stratum	Strata used in the survey design
panel	Identifies base sample by panel name and Oversample by OverSamp
auxiliary variables	Remaining columns are from the sample frame provided

###  __Projection Information__

Projected Coordinate System: FDEP Albers HARN

Projection: Albers

False_Easting: 400000.00000000

False_Northing: 0.00000000

Central_Meridian: -84.00000000

Standard_Parallel_1: 24.00000000

Standard_Parallel_2: 31.50000000

Latitude_Of_Origin: 24.00000000

Linear Unit: Meter


Geographic Coordinate System: GCS_North_American_1983_HARN

Datum: D_North_American_1983_HARN

Prime Meridian: Greenwich

Angular Unit: Degree



### __Evaluation Process__
The survey design weights that are given in the design file assume that the survey design is implemented as designed.  That is, only the sites that are in the base sample (not in the over sample) are used, and all the base sites are used.  This may not occur due to (1)  sites not being a member of the target population, (2) landowners deny access to a site, (3) a site is physically inaccessible (safety reasons), or (4) site not sampled for other reasons.  Typically, users prefer to replace sites that can not be sampled with other sites to achieve the sample size planned.  The site replacement process is described above.  When sites are replaced, the survey design weights are no longer correct and must be adjusted.  The weight adjustment requires knowing what happened to each site in the base design and the over sample sites.  EvalStatus is initially set to “NotEval” to indicate that the site has yet to be evaluated for sampling.  When a site is evaluated for sampling, then the EvalStatus for the site must be changed.  Recommended codes are:

EvalStatus Code	 |    Name	      |       Meaning

TS	             |    Target Sampled	 |  site is a member of the target population and was sampled

LD	             |    Landowner Denial	| landowner denied access to the site

PB	              |   Physical Barrier|	physical barrier prevented access to the site

NT	              |   Non-Target	    |  site is not a member of the target population

NN	               |  Not Needed	    |  site is a member of the over sample and was not evaluated for sampling



Other codes		Many times useful to have other codes.  For example, rather than use NT, may use specific codes indicating why the site was non-target.

### Statistical Analysis
Any statistical analysis of data must incorporate information about the monitoring survey design.  In particular, when estimates of characteristics for the entire target population are computed, the statistical analysis must account for any stratification or unequal probability selection in the design.  Procedures for doing this are available from the Aquatic Resource Monitoring web page given in the bibliography.  A statistical analysis library of functions is available from the web page to do common population estimates in the statistical software environment R. 

Questions and Inquiries about site selections

Chris Sedlacek

Environmental Consultant, Watershed Monitoring 

E-Mail: christopher.sedlacek@dep.state.fl.us




Questions and Inquiries about GRTS sampling program/protocols 

Anthony (Tony) R. Olsen

USEPA NHEERL

Western Ecology Division

200 S.W. 35th Street

Corvallis, OR 97333

Voice: (541) 754-4790

Fax: (541) 754-4716

email: Olsen.Tony@epa.gov

### __Bibliography:__

Diaz-Ramos, S., D. L. Stevens, Jr, and A. R. Olsen. 1996. EMAP Statistical Methods Manual. EPA/620/R-96/002, U.S. Environmental Protection Agency, Office of Research and Development, NHEERL-Western Ecology Division, Corvallis, Oregon.

Stevens, D.L., Jr. 1997. Variable density grid-based sampling designs for continuous spatial populations. Environmetrics, 8:167-95.

Stevens, D.L., Jr. and Olsen, A.R. 1999. Spatially restricted surveys over time for aquatic resources. Journal of Agricultural, Biological, and Environmental
Statistics, 4:415-428

Stevens, D. L., Jr., and A. R. Olsen. 2003. Variance estimation for spatially balanced samples of environmental resources. Environmetrics 14:593-610.

Stevens, D. L., Jr., and A. R. Olsen. 2004. Spatially-balanced sampling of natural resources in the presence of frame imperfections. Journal of American Statistical Association:99:262-278.


Web Page: http://www.epa.gov/nheerl/arm

